---
title: "Package Architecture"
subtitle: "Design, golden fixtures, and dataset catalog for TemporalHazard"
author: "John Ehrlinger"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
vignette: >
  %\VignetteIndexEntry{Package Architecture}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
has_pkg <- requireNamespace("TemporalHazard", quietly = TRUE)
if (has_pkg) library(TemporalHazard)
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
                      eval = has_pkg)
```

# Overview

TemporalHazard is a pure-R reimplementation of the C/SAS HAZARD procedure
originally developed at the University of Alabama at Birmingham (UAB) for
multi-phase parametric hazard modeling (Blackstone, Naftel, and Turner 1986).
The SAS/C code and this R package are currently developed and maintained at
The Cleveland Clinic Foundation, and the R code was wholly developed at The
Cleveland Clinic Foundation. The package provides a
unified framework for fitting additive hazard models with an arbitrary
number of temporal phases. The early (`"cdf"`) and `"hazard"` phases use
the three-parameter `decompos(t; t_half, nu, m)` family, and the late
(`"g3"`) phase its own four-parameter shape. The generalized temporal
decomposition extends naturally to longitudinal mixed-effects settings
(Rajeswaran et al. 2018).

This vignette documents the internal architecture: how source files are
organized, how functions compose into the fitting pipeline, how golden
fixtures catch regressions, and what reference datasets
ship with the package.


# Source file organization

The R source lives in `R/` and follows a layered design. Lower layers
know nothing about higher layers; control flows downward from the API
through distribution-specific optimizers to shared numerical primitives.

```{r file-layout, echo=FALSE}
layout <- data.frame(
  Layer = c(
    rep("User API", 2),
    rep("Multiphase engine", 3),
    rep("Single-phase likelihoods", 4),
    rep("Shared infrastructure", 4)
  ),
  File = c(
    "hazard_api.R", "argument_mapping.R",
    "likelihood-multiphase.R", "phase-spec.R", "decomposition.R",
    "likelihood-weibull.R", "likelihood-exponential.R",
    "likelihood-loglogistic.R", "likelihood-lognormal.R",
    "optimizer.R", "math_primitives.R", "formula-helpers.R",
    "parity-helpers.R"
  ),
  Purpose = c(
    "hazard(), predict.hazard(), print/summary/coef/vcov S3 methods",
    "SAS HAZARD -> R parameter translation table",
    "Additive N-phase likelihood, cumhaz/hazard evaluators, multi-start optimizer",
    "hzr_phase() constructor, validators, starting-value assembly",
    "Unified decompos(t; t_half, nu, m) parametric family",
    "Weibull PH likelihood, gradient, optimizer",
    "Exponential (constant hazard) likelihood",
    "Log-logistic (proportional odds) likelihood",
    "Log-normal (AFT) likelihood",
    "Generic L-BFGS-B/BFGS optimizer with Hessian-based vcov",
    "Numerically stable log1pexp, log1mexp, clamp_prob",
    "Surv() formula parsing for right/left/interval censoring",
    "Stubs for cross-validating against C HAZARD binary"
  )
)
knitr::kable(layout, caption = "Selected R source files by architectural layer")
```


# Function call graph

The primary user entry point is `hazard()`, which dispatches to
distribution-specific optimizers. The diagram below shows the call flow
for a multiphase fit.

```
Call graph for a multiphase hazard fit
--------------------------------------

hazard(dist='multiphase')
+-- .hzr_optim_multiphase()
|   +-- Resolve per-phase design matrices
|   +-- .hzr_phase_start() per phase
|   +-- .hzr_optim_generic()
|       +-- stats::optim(method='BFGS')
|           +-- .hzr_logl_multiphase()
|               +-- .hzr_split_theta()
|               +-- .hzr_multiphase_cumhaz()
|               |   +-- hzr_phase_cumhaz() / hzr_decompos_g3()
|               +-- .hzr_multiphase_hazard()
|                   +-- hzr_phase_hazard() / hzr_decompos_g3()
+-- predict.hazard()
    +-- .hzr_multiphase_cumhaz()
    +-- .hzr_multiphase_hazard()
```

For single-phase distributions (Weibull, exponential, log-logistic,
log-normal), `hazard()` dispatches to the corresponding
`.hzr_optim_<dist>()` function, which calls `.hzr_optim_generic()` with
the distribution-specific log-likelihood and gradient.


## Key internal functions

### Theta vector layout

The multiphase model concatenates per-phase sub-vectors into a single
flat `theta` on the **internal (estimation) scale**:

```{r theta-layout, echo=FALSE}
theta_layout <- data.frame(
  `Phase type` = c("cdf / hazard", "g3", "constant"),
  `Sub-vector` = c(
    "[log_mu, log_t_half, nu, m, beta_1, ..., beta_p]",
    "[log_mu, log_tau, gamma, alpha, eta, beta_1, ..., beta_p]",
    "[log_mu, beta_1, ..., beta_p]"
  ),
  `Count` = c("4 + p", "5 + p", "1 + p")
)
knitr::kable(theta_layout, caption = "Internal theta layout per phase")
```

`.hzr_split_theta()` partitions the concatenated vector into a named
list of per-phase sub-vectors. `.hzr_unpack_phase_theta()` extracts
named parameters (`log_mu`, `log_t_half`, `nu`, `m`, `beta`) from each
sub-vector.

### Phase specification

`hzr_phase(type, t_half, nu, m, formula)` creates a lightweight S3
object that stores the phase type, initial shape parameters, and an
optional phase-specific covariate formula. The constructor validates
inputs and returns `NA` for shape parameters of constant phases.

A list of `hzr_phase` objects is validated by `.hzr_validate_phases()`,
which auto-names unnamed phases ("phase_1", "phase_2", ...) and catches
the common mistake of passing a bare `hzr_phase` instead of a list.

### Decomposition engine

`hzr_decompos(time, t_half, nu, m)` is the mathematical core. It
computes `G(t)` (CDF), `g(t)` (density), and `h(t)` (hazard) for the
three-parameter family across six valid sign combinations of `nu` and
`m`. The phase-level helpers `hzr_phase_cumhaz()` and
`hzr_phase_hazard()` wrap `hzr_decompos()` and apply the phase type
mapping:

| Phase type | $\Phi(t)$ | $\phi(t)$ |
|:-----------|:----------|:----------|
| `"cdf"`    | $G_1(t)$    | $g_1(t)$    |
| `"hazard"` | $-\log(1-G_1(t))$ | $h_1(t)$ |
| `"g3"`     | $G_3(t; \tau, \gamma, \alpha, \eta)$ | $g_3(t)$ |
| `"constant"` | $t$     | $1$       |

### Multi-start optimization

`.hzr_optim_multiphase()` runs a multi-start strategy: the first start
uses the assembled starting values (from `theta` or `hzr_phase` specs);
subsequent starts perturb randomly (sd = 0.5). The best log-likelihood
across all starts is kept. This helps the optimizer escape the shallow local
optima that are common in multiphase models.

The optimizer delegates to `.hzr_optim_generic()`, which wraps
`stats::optim()` with method `"BFGS"` (unconstrained; all scale
parameters are log-transformed) and, when a BFGS stop fails SAS/C
HAZARD's relative-gradient test, continues with `stats::nlm()`. The Hessian at the converged point,
inverted for the variance-covariance matrix, is assembled analytically
when every row is an exact event or right-censored (counting-process start
times included): `.hzr_hessian_multiphase()` for multiphase fits, whose
phase-shape second derivatives inside it are finite differences, and a
closed-form Hessian for each single-phase distribution. When any row is
left- or interval-censored, the analytic Hessian declines and
`numDeriv::hessian()` computes it numerically. `numDeriv` is a Suggests,
so without it those fits have no standard errors.


# Golden fixture system

Golden fixtures are pre-fitted model results saved as `.rds` files in
`inst/fixtures/`. They are **regression anchors**: each test run
refits the model on the same data and compares estimates to the stored
values. This catches regressions when the likelihood, gradient, or
optimizer changes.

## Fixture format

Every fixture is a named list with a consistent structure:

```r
list(
  description = "Human-readable description",
  data = list(time, status, [x], n, events),
  seed = 42,
  true_params = list(...),   # simulation truth (synthetic fixtures)
  fit = list(
    theta     = <named numeric>,  # converged parameter vector
    logl      = <scalar>,         # log-likelihood at convergence
    converged = <logical>,        # convergence flag
    vcov      = <matrix>          # asymptotic variance-covariance
  ),
  timestamp = <POSIXct>
)
```

For C-reference fixtures (e.g., `mp_c_reference_kul.rds`), the `fit`
element is replaced by `c_reference` containing the C binary output:
parameter estimates, standard errors, variance-covariance matrix, and
the C log-likelihood.

## Current fixtures

```{r fixtures, echo=FALSE}
fixtures <- data.frame(
  File = c(
    "hz_univariate.rds",
    "hm_multivariate.rds",
    "hm_edge_case.rds",
    "hz_loglogistic.rds",
    "hz_lognormal.rds",
    "mp_synthetic_3phase.rds",
    "mp_c_reference_kul.rds"
  ),
  Distribution = c(
    "Weibull", "Weibull", "Weibull",
    "Log-logistic", "Log-normal",
    "Multiphase (3)", "Multiphase (3)"
  ),
  Description = c(
    "Univariable shape estimation (n=100)",
    "2 covariates (n=100)",
    "Edge case: n=20, 3 covariates",
    "Univariable (n=80)",
    "Univariable (n=80)",
    "Synthetic early CDF + constant + late hazard (n=500)",
    "KUL CABG dataset + C HAZARD binary reference output (n=5880)"
  ),
  Source = c(
    rep("Synthetic (seed=42)", 5),
    "Synthetic (seed=42)",
    "Clinical + C binary"
  )
)
knitr::kable(fixtures, caption = "Golden fixture inventory")
```

## Regenerating fixtures

Fixtures must be regenerated whenever the model parameterization changes.
The generators are maintainer-only helpers kept in `data-raw/` (outside the
installed package). Run them interactively from the package source tree
after `devtools::load_all()` and sourcing the generator script:

```r
devtools::load_all()
source("data-raw/golden_fixtures.R")

# Each generator requires an explicit output directory (no default path
# is written). Use the project source dir to persist, or tempdir() to test.
dir <- "inst/fixtures"

# Single-phase distributions
.hzr_create_synthetic_golden_fixtures(dir, seed = 42L)   # Weibull variants
.hzr_create_loglogistic_golden_fixture(dir, seed = 42L)  # Log-logistic
.hzr_create_lognormal_golden_fixture(dir, seed = 42L)    # Log-normal

# Multiphase
.hzr_create_multiphase_golden_fixture(dir, seed = 42L)   # Synthetic 3-phase
.hzr_create_c_reference_kul_fixture(dir)                  # C binary reference
```

Pass an explicit `seed` for reproducibility; the global RNG state is saved
and restored, so the caller's stream is unaffected. With the default
(`seed = NULL`) no seed is set. Commit the updated `.rds` files alongside
any code changes.


# Testing strategy

The test suite (`tests/testthat/`) is organized into four tiers:

```{r test-tiers, echo=FALSE}
tiers <- data.frame(
  Tier = c(
    "Unit tests",
    "Distribution tests",
    "Integration tests",
    "Parity tests"
  ),
  Files = c(
    "test-math-primitives, test-decomposition, test-phase-spec, test-argument-mapping",
    paste("test-gradient-weibull, test-exponential-dist,",
          "test-loglogistic-dist, test-lognormal-dist, test-multiphase-gradient"),
    paste("test-hazard-api, test-predict-types, test-interval-censoring-*,",
          "test-time-varying-*, test-multiphase-likelihood, test-multiphase-api"),
    "test-parity-core, test-parity-edge-cases, test-parity-c-binary, test-multiphase-parity"
  ),
  Purpose = c(
    "Verify individual functions in isolation",
    "Likelihood, gradient, and optimizer for each distribution",
    "End-to-end: hazard() -> predict() -> summary() pipeline, censoring types, multiphase wiring",
    "Golden fixture round-trip, C binary cross-validation"
  )
)
knitr::kable(tiers, caption = "Test suite tiers")
```

### Multiphase parity tests

The multiphase parity tests (`test-multiphase-parity.R`) validate
against the C HAZARD binary output for the KUL CABG dataset:

1. **Likelihood evaluation**: evaluates the R log-likelihood at the C
   converged parameters and asserts it matches the C output (-3740.52).

2. **Decomposition consistency**: verifies phase additivity and CDF
   saturation at the C reference parameter values.

3. **Conservation of events**: checks that the model-implied expected
   events ($\sum [1 - \exp(-H(t_i))]$) match the observed event count
   (545), as reported by the C binary (544.9993).

4. **Profile standard errors**: computes a numerical Hessian varying
   only the 3 log(mu) parameters (shapes held fixed), matching the C
   binary's estimation strategy, and compares standard errors.

5. **Full fit convergence**: fits the R multiphase optimizer on the
   full dataset with informed starting values and checks that the
   log-likelihood meets or exceeds the C reference.


# Dataset catalog {#sec-datasets}

TemporalHazard ships five clinical reference datasets in
`inst/extdata/`, converted from the original C/SAS HAZARD test data.
These datasets are used in vignette examples and parity testing.

```{r datasets, echo=FALSE}
datasets <- data.frame(
  File = c("avc.csv", "cabgkul.csv", "omc.csv", "tga.csv", "valves.csv"),
  Study = c(
    "Atrioventricular canal repair",
    "Coronary artery bypass grafting (KU Leuven)",
    "Open mitral commissurotomy",
    "Transposition of great arteries (arterial switch)",
    "Primary valve replacement"
  ),
  n = c(310, 5880, 339, 470, 1533),
  Events = c("70 deaths", "545 deaths", "thromboembolic events (repeated)", "deaths", "deaths, PVE, reoperation"),
  Covariates = c(
    "NYHA, age, anatomy, era",
    "None (intercept only)",
    "TE events (repeated measures)",
    "anatomy, coronary pattern, era",
    "age, NYHA, valve position, pathology, race"
  ),
  `SAS origin` = c(
    "hz.death.AVC, hm.death.AVC",
    "hz.deadp.KUL",
    "hz.te123.OMC",
    "hs.dthar.TGA",
    "hm.deadp.VALVES"
  )
)
knitr::kable(datasets, caption = "Reference datasets (lazy-loaded via `data()`)")
```


## Loading datasets

```{r load-demo}
# Datasets are lazy-loaded with the package — just reference them directly.
# Raw CSVs are also available in inst/extdata/ for advanced use:
#   read.csv(system.file("extdata", "cabgkul.csv", package = "TemporalHazard"))

data(cabgkul)
str(cabgkul)

data(avc)
str(avc)
```


## Dataset details

### AVC (atrioventricular canal repair)

The AVC dataset contains 310 patients who underwent repair of
atrioventricular septal defects. It is the primary example dataset used
throughout the hazard modeling examples and has the richest set of
covariates for multivariable analysis.

```{r avc-detail, echo=FALSE}
avc_vars <- data.frame(
  Variable = c("study", "status", "inc_surg", "opmos", "age", "mal",
               "com_iv", "orifice", "dead", "int_dead", "op_age"),
  Label = c("Study number", "NYHA functional class (I-IV)", "Surgical grade of AV valve incompetence",
            "Date of operation (months since Jan 1967)", "Age (months) at repair",
            "Important associated cardiac anomaly", "Interventricular communication",
            "Accessory left AV valve orifice", "Death (event indicator)",
            "Follow-up interval (months)", "Interaction: opmos x age"),
  Type = c("character", rep("numeric", 10))
)
knitr::kable(avc_vars, caption = "AVC dataset variables")
```

### CABG/KUL (coronary artery bypass grafting)

The KUL dataset is a large series of 5880 primary isolated CABG patients
from KU Leuven (1971--July 1987). It is the primary benchmark for
C binary parity testing because it has the simplest structure
(intercept-only, right-censored) combined with a large sample size that
exercises all three temporal phases.

The original SAS analysis (`hz.deadp.KUL.sas`) also includes
return-of-angina and reintervention endpoints (recorded in the original
fixed-width file with 6 columns), but only the death endpoint
(`int_dead`, `dead`) is extracted for parity testing.

### OMC (open mitral commissurotomy)

The OMC dataset contains 339 patients and is unique in the collection
because it involves **repeated thromboembolic events** (up to 3 per
patient). The SAS analysis transforms the dataset into a
repeated-events format in which each interval starts at the previous
event (STARTTME) and ends at the next event or censoring (CENSORED).
The SAS job names STARTTME in its `LCENSOR` statement, and the HAZARD
documentation calls that left censoring, but it records delayed entry
(left truncation). The analysis therefore exercises the counting-process
start-time term of the likelihood, not the left- or interval-censoring
terms.

### TGA (transposition of great arteries)

The TGA dataset contains 470 patients who underwent the arterial switch
operation. It includes derived variables (logarithmic and inverse
transforms of clinical measurements) and is primarily used for
sensitivity analysis and internal validation examples.

### Valves (primary valve replacement)

The VALVES dataset is the largest multivariable example with 1533
patients and multiple endpoints (death, prosthetic valve endocarditis,
bioprosthesis degeneration, reoperation). The SAS analysis
(`hm.deadp.VALVES.sas`) fits a multivariable 3-phase model with 10
covariates across multiple event types.


# SAS/C parameter mapping

The original C/SAS HAZARD procedure uses a different parameterization
than TemporalHazard. The `hzr_argument_mapping()` function documents
the full translation.

```{r mapping}
mapping <- hzr_argument_mapping()
knitr::kable(
  mapping[, c("legacy_input", "r_parameter", "implementation_status", "notes")],
  caption = "SAS HAZARD to R parameter mapping (excerpt)"
)
```


## Early phase (G1) mapping

The SAS early phase uses four parameters: DELTA, RHO (or THALF), NU, M.
With DELTA = 0 they reduce to the three-parameter decompos family:

- **DELTA**: time transformation `B(t) = (exp(delta*t) - 1)/delta`.
  When DELTA = 0 (the common case), `B(t) = t` and the transformation
  drops out. Non-zero DELTA is not supported.
- **RHO / THALF**: scale parameter. THALF maps directly to `t_half` in
  `hzr_phase()`. RHO has no argument of its own in R: it is fixed by NU,
  THALF and M, and for M > 0 and NU > 0 it is
  `RHO = NU * THALF * ((2^M - 1)/M)^NU`.
- **NU**: time exponent. Maps directly.
- **M**: shape exponent. Maps directly.

## Late phase (G3) mapping

The SAS late phase uses the G3 decomposition with four parameters, now
directly supported via `hzr_phase("g3", ...)`:

- **TAU** --> `tau` (scale parameter)
- **GAMMA** --> `gamma` (time exponent)
- **ALPHA** --> `alpha` (shape parameter; `alpha = 0` gives exponential case)
- **ETA** --> `eta` (outer exponent)

The G3 formula (for `alpha > 0`) is:
$G_3(t) = \left(\left((t/\tau)^\gamma + 1\right)^{1/\alpha} - 1\right)^\eta$

Unlike G1, G3 is unbounded: it can grow without limit, making it
suitable for late-phase rising hazards. For the KUL benchmark with
`tau = 1, gamma = 3, alpha = 1, eta = 1`, this simplifies to $G_3(t) = t^3$.


# Version history

Versions are plain three-part numbers (major.minor.patch). The early
milestones were:

- **v0.1.0** (single-phase engine): Weibull, exponential, log-logistic,
  log-normal distributions with formula interface, predict, and golden
  fixture testing.
- **v0.9.0** (multiphase engine): N-phase additive cumulative hazard,
  `hzr_phase()` specification, decomposition engine, C binary parity
  tests, dataset catalog.
- **v0.9.1**: vignette suite, roxygen multiphase examples,
  CI workflow fixes (load_pkgload), SAS missing-value handling
  (`na.strings`), print.hazard phase labels, and README refresh.

Every release since, up to the installed version, is listed in the
package NEWS: `news(package = "TemporalHazard")`.


# References {.unnumbered}

Blackstone EH, Naftel DC, Turner ME Jr. The decomposition of time-varying
hazard into phases, each incorporating a separate stream of concomitant
information. *J Am Stat Assoc.* 1986;81(395):615-624.
doi: [10.1080/01621459.1986.10478314](https://doi.org/10.1080/01621459.1986.10478314)

Rajeswaran J, Blackstone EH, Ehrlinger J, Li L, Ishwaran H, Parides MK.
Probability of atrial fibrillation after ablation: Using a parametric
nonlinear temporal decomposition mixed effects model. *Stat Methods Med Res.*
2018;27(1):126-141.
doi: [10.1177/0962280215623583](https://doi.org/10.1177/0962280215623583)
