---
title: "Fitting Hazard Models"
subtitle: "From intercept-only to multiphase additive hazard"
format:
  html:
    code-fold: false
    toc: true
    toc-depth: 3
    number-sections: true
vignette: >
  %\VignetteIndexEntry{Fitting Hazard Models}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
has_pkg <- requireNamespace("TemporalHazard", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_pkg
)
```

```{r}
#| label: setup
library(TemporalHazard)
library(survival)
library(ggplot2)
```

This vignette walks the core model-fitting workflow from the inside
out: intercept-only fits to establish the baseline hazard shape, then
covariates on top, then the multiphase decomposition, then
multi-endpoint analyses on the same cohort. Every example uses a
clinical dataset shipped with the package. If you haven't seen the
basics — what a parametric hazard model is, why we use the
`Surv(time, status)` formula — start with `vignette("getting-started")`
first; this vignette assumes that context.

The progression matters. Single-distribution intercept-only fits tell
you whether the baseline hazard shape is monotone (Weibull territory)
or has structure that demands a multiphase decomposition.
Multivariable fits add covariate effects on top of a shape you already
trust. Multiphase fits split the baseline shape into clinically
interpretable phases. Multi-endpoint analyses reuse all the above for
separate clinical outcomes — death, reoperation, infection — on the
same patient cohort.


## Intercept-only model: CABG survival (KU Leuven)

The `cabgkul` dataset contains 5,880 patients who underwent primary isolated
coronary artery bypass grafting at KU Leuven between 1971 and 1987.  With only
two columns --- follow-up time and death indicator --- it is the simplest
starting point.

```{r}
#| label: kul-data
data(cabgkul)
str(cabgkul)
```

Fit an intercept-only Weibull. With no covariates on the right-hand
side of the formula the model estimates only the baseline hazard
shape — the scale `mu` and exponent `nu` of a Weibull curve fit to all
5,880 patients pooled. This is the right starting point for any new
dataset: before asking which covariates matter, ask whether a single
monotone hazard even fits the population-level pattern.

```{r}
#| label: kul-weibull
fit_kul <- hazard(
  Surv(int_dead, dead) ~ 1,
  data  = cabgkul,
  dist  = "weibull",
  theta = c(mu = 0.10, nu = 1.0),
  fit   = TRUE
)

fit_kul
```

The summary tells us where the optimizer landed; the picture tells us
whether that landing point matches the data. Plot the fitted survival
curve on a fine time grid and overlay the Kaplan-Meier step function
from the raw cohort.

```{r}
#| label: fig-kul-km
#| fig-cap: "Weibull parametric survival vs. Kaplan-Meier (CABG, KU Leuven)"
#| fig-width: 7
#| fig-height: 4.5
t_grid <- seq(0.01, max(cabgkul$int_dead) * 0.9, length.out = 200)
nd     <- data.frame(time = t_grid)
surv   <- predict(fit_kul, newdata = nd, type = "survival") * 100

km     <- survfit(Surv(int_dead, dead) ~ 1, data = cabgkul)
km_df  <- data.frame(time = km$time, survival = km$surv * 100)

ggplot() +
  geom_step(data = km_df, aes(time, survival, colour = "Kaplan-Meier"),
            linewidth = 0.5) +
  geom_line(data = data.frame(time = t_grid, survival = surv),
            aes(time, survival, colour = "Weibull"), linewidth = 1) +
  scale_colour_manual(values = c("Weibull" = "#0072B2",
                                 "Kaplan-Meier" = "#D55E00")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(x = "Months after CABG", y = "Freedom from death (%)",
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

A single Weibull captures the broad trend but misses the distinct early
operative risk and late attrition that the KM curve reveals.  This motivates
the multiphase approach below.


## Multivariable model: AVC repair

The `avc` dataset has 310 patients who underwent atrioventricular canal
repair, with 9 candidate covariates spanning patient demographics (age,
NYHA status), anatomical features (malalignment, orifice morphology),
intra-operative grading (`inc_surg` = surgical grade of AV valve
incompetence), and post-operative complications (`com_iv` = grade IV
complications). We drop incomplete rows so the design matrix is
rectangular, then look at the column types and ranges.

```{r}
#| label: avc-data
data(avc)
avc <- na.omit(avc)
str(avc)
```

Now we put covariates on the right-hand side of the formula and refit.
The `theta` vector grows: two Weibull shape parameters (`mu`, `nu`) plus
six covariate coefficients (`beta1`..`beta6`), each starting at zero.
The optimizer estimates a log-hazard-ratio for every covariate jointly
with the Weibull shape — so the shape and the covariate effects are
identified from the same likelihood, not sequentially.

```{r}
#| label: avc-mv
fit_avc <- hazard(
  Surv(int_dead, dead) ~ age + status + mal + com_iv + inc_surg + orifice,
  data  = avc,
  dist  = "weibull",
  theta = c(mu = 0.20, nu = 1.0, rep(0, 6)),
  fit   = TRUE,
  control = list(maxit = 500)
)

fit_avc
```

Each coefficient is a log-hazard-ratio: positive means higher risk,
negative means lower, zero means no effect. The large positive
coefficients on `mal` (anatomical malalignment) and `com_iv`
(grade IV post-operative complications) flag these as the dominant
risk markers in this cohort. The standard errors and Wald
z-statistics in the summary tell you which effects are well
identified and which are noise — a coefficient with a z-statistic
near zero contributes essentially nothing the data can defend.


## Multiphase model: additive hazard decomposition

The single-Weibull fits above gave us a curve that's mediocre
everywhere instead of right anywhere. That's a structural limitation
of a monotone parametric shape, not something more iterations will
fix. The Blackstone–Naftel–Turner framework's key idea is to split
the hazard into a *sum* of phase-specific contributions, each with its
own temporal shape and its own scale:

$$H(t \mid x) = \sum_{j=1}^{J} \mu_j(x) \cdot \Phi_j(t)$$

Each $\Phi_j(t)$ is a phase-specific unit-scaled curve (early-peaking
saturating, flat constant, late-rising polynomial) and each $\mu_j(x)$
is the phase-specific scale, possibly modulated by covariates. The
phases overlap and add — no switching, no thresholds — so the total
instantaneous hazard at any $t$ is the sum of the per-phase rates.
See `vignette("getting-started")` for the longer-form motivation; what
follows here is the practical workflow for *fitting* one.

For AVC we'll use two phases — an early phase to absorb the
operative-window mortality, and a constant phase for the background
rate. AVC patients don't have a clear late-deterioration regime over
this follow-up window, so a third (g3) phase would be unidentified.
We fix the shape parameters and estimate only the scales, matching
the workflow you'd run against a SAS HAZARD reference fit.

```{r}
#| label: avc-mp
fit_mp <- hazard(
  Surv(int_dead, dead) ~ 1,
  data   = avc,
  dist   = "multiphase",
  phases = list(
    early    = hzr_phase("cdf", t_half = 0.5, nu = 1, m = 1,
                          fixed = "shapes"),
    constant = hzr_phase("constant")
  ),
  fit     = TRUE,
  control = list(n_starts = 5, maxit = 1000)
)

summary(fit_mp)
```

The diagnostic that matters is whether the multiphase fit actually
out-performs the single Weibull against the data. Plot both parametric
curves against the same Kaplan-Meier reference so we can see, by eye,
where each model is honest and where each is reaching.

```{r}
#| label: fig-avc-compare
#| fig-cap: "Single-phase Weibull vs. multiphase model against Kaplan-Meier (AVC)"
#| fig-width: 7
#| fig-height: 4.5
t_grid <- seq(0.01, max(avc$int_dead) * 0.95, length.out = 200)
nd     <- data.frame(time = t_grid)

km_avc <- survfit(Surv(int_dead, dead) ~ 1, data = avc)
km_df  <- data.frame(time = km_avc$time, survival = km_avc$surv * 100)

fit_wb <- hazard(
  Surv(int_dead, dead) ~ 1, data = avc, dist = "weibull",
  theta = c(mu = 0.20, nu = 1.0), fit = TRUE
)

surv_wb <- predict(fit_wb, newdata = nd, type = "survival") * 100
surv_mp <- predict(fit_mp, newdata = nd, type = "survival") * 100

plot_df <- rbind(
  data.frame(time = t_grid, survival = surv_wb, Model = "Single Weibull"),
  data.frame(time = t_grid, survival = surv_mp, Model = "Multiphase (2-phase)")
)

ggplot() +
  geom_step(data = km_df, aes(time, survival), colour = "grey50",
            linewidth = 0.5) +
  geom_line(data = plot_df, aes(time, survival, colour = Model),
            linewidth = 1) +
  scale_colour_manual(values = c("Single Weibull" = "#E69F00",
                                 "Multiphase (2-phase)" = "#0072B2")) +
  scale_y_continuous(limits = c(0, 100)) +
  annotate("text", x = max(t_grid) * 0.6, y = 95, label = "KM (grey)",
           size = 3, colour = "grey50") +
  labs(x = "Months after AVC repair", y = "Freedom from death (%)",
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The multiphase model tracks the KM curve much more closely than the
single Weibull, especially across the steep early-mortality window —
which is exactly where the single Weibull was forced to compromise.
The constant phase then carries the slow post-recovery attrition. The
point isn't that multiphase always wins; it's that *when the data has
phase structure*, fitting that structure explicitly is strictly more
honest than averaging it away into one monotone curve.


## Multi-endpoint models: heart valve replacement

The `valves` dataset (1,533 patients) has multiple time-to-event endpoints ---
death, prosthetic valve endocarditis (PVE), and reoperation --- each with its
own follow-up time and event indicator.  The same `hazard()` call fits each
endpoint independently:

Start with the death endpoint. We use age at operation, NYHA class,
and mechanical-valve indicator as covariates — the clinically
canonical set for survival after valve replacement.

```{r}
#| label: valves-death
data(valves)
valves <- na.omit(valves)

fit_death <- hazard(
  Surv(int_dead, dead) ~ age_cop + nyha + mechvalv,
  data  = valves,
  dist  = "weibull",
  theta = c(mu = 0.10, nu = 1.0, rep(0, 3)),
  fit   = TRUE,
  control = list(maxit = 500)
)

fit_death
```

Switch endpoints. Same data, same package, but now we model time to
prosthetic valve endocarditis instead of death. The covariate list
shifts to match the clinical question: `nve` (native-valve endocarditis
history) replaces `nyha` because functional class is less informative
for infection risk than prior endocarditis exposure. The fit returns
its own MLE, coefficients, and standard errors completely independent
of the death model.

```{r}
#| label: valves-pve
fit_pve <- hazard(
  Surv(int_pve, pve) ~ age_cop + nve + mechvalv,
  data  = valves,
  dist  = "weibull",
  theta = c(mu = 0.02, nu = 1.0, rep(0, 3)),
  fit   = TRUE,
  control = list(maxit = 500)
)

fit_pve
```

Each endpoint gets its own model with its own covariates, but the
hazard model structure — temporal shape plus covariate effects — stays
the same whatever the clinical endpoint. Repeating the workflow for a
third endpoint (reoperation, for example) is mechanical: swap the
`Surv(...)` columns, swap the covariates, refit. The advantage over
running three separate analyses in different tools is that the
predictions, diagnostics, and uncertainty quantification all come from
the same package — there's no risk of subtle differences in censoring
handling or estimator choice between endpoints.


## Interval and left censoring

Right censoring is by far the most common censoring type in clinical
survival data — a patient is still alive at last follow-up, so all we
know is that their event time exceeds the observed window. Two other
types arise frequently enough to warrant explicit handling.

**Interval censoring** occurs when the event is known to have happened
*between* two observation times, but not at a specific time.  The
canonical example is clinic-visit data: a patient is confirmed alive at
month 12 and found dead at month 24, so the event occurred somewhere in
$(12, 24)$.  A naive analysis treats the event as right-censored at 12
or exactly observed at 24; both introduce bias.

**Left censoring** is the mirror: the event occurred *before* the first
observation time (time_upper), so it was already established at the
moment of first contact — $T \leq t_{\text{upper}}$.

### Status codes

The package encodes censoring type in the `status` vector:

| Status | Type | Interpretation |
|-------:|------|----------------|
| `1` | Exact event | Observed at `time` |
| `0` | Right-censored | Event did not occur by `time` |
| `2` | Interval-censored | Event occurred in `(time_lower, time_upper)` |
| `-1` | Left-censored | Event occurred before `time` (upper bound) |

For interval-censored rows, supply both `time_lower` and `time_upper`.  You
can pass them as direct arguments to `hazard()`, or hand it a `Surv()`
object instead: `Surv(lower, upper, event, type = "interval")` and
`Surv(lower, upper, type = "interval2")` both work, as do
`Surv(time, event, type = "left")` for left censoring and
`Surv(start, stop, event)` for counting-process (repeating-event) data.

One thing to watch, because the two interfaces do not use the same numbers.
The table above is this package's coding, and it is what you pass to the
`status` argument.  `Surv()` has its own scheme --- under
`type = "interval"` it reads `0` as right-censored, `1` as an exact event,
`2` as left-censored and `3` as interval-censored --- and `hazard()`
translates it for you when it parses the formula.  So use whichever coding
belongs to the interface you are writing in, and don't carry one set of
codes across to the other.

### Mixed censoring example

We simulate 400 cardiac surgery patients observed at semi-annual clinic
visits (every 6 months, 4--8 scheduled visits per patient).  If the
event falls between two consecutive visits the observation is
interval-censored; if it falls after the last visit the patient is
right-censored.

```{r}
#| label: interval-simulate
set.seed(101)
n         <- 400
visit_gap <- 6   # months between visits

# True event times: Weibull mu = 0.025, nu = 1.0 (median ~28 months)
true_time <- rexp(n, rate = 0.025)  # nu=1 → exponential

n_visits  <- sample(4:8, n, replace = TRUE)

status     <- integer(n)
time       <- numeric(n)
time_lower <- numeric(n)
time_upper <- numeric(n)

for (i in seq_len(n)) {
  visits <- seq(visit_gap, n_visits[i] * visit_gap, by = visit_gap)
  t_ev   <- true_time[i]
  brk    <- findInterval(t_ev, c(0, visits))  # which interval contains t_ev

  if (brk > length(visits)) {
    # Event after last visit: right-censored.
    # time_lower = 0: with H(time_lower) = H(0) = 0, the right-censored
    # contribution reduces to -(H(time) - H(0)) = -H(time), which is correct.
    status[i]     <- 0L
    time[i]       <- max(visits)
    time_lower[i] <- 0
    time_upper[i] <- max(visits)
  } else {
    # Event within the visit window: interval-censored in (lower, upper)
    lower_bound   <- if (brk == 1L) 0 else visits[brk - 1L]
    upper_bound   <- visits[brk]
    status[i]     <- 2L
    time[i]       <- lower_bound
    time_lower[i] <- lower_bound
    time_upper[i] <- upper_bound
  }
}

table(status)
```

::: {.callout-note}
**`time_lower` as a counting-process entry time.**  In the Weibull and
multiphase likelihoods, `time_lower` doubles as the *entry* (left-truncation)
time for right-censored and exact-event rows (`status %in% c(0, 1)`): when
`0 < time_lower < time`, the row contributes H(stop) − H(start), the
counting-process form used for epoch-decomposed repeated events.  For an
ordinary right-censored or event row with no entry time, leave `time_lower`
at `0` or omit it.  Setting `time_lower = time` is treated as *no entry* —
`time_lower` acts as an entry time only when strictly less than `time`, so
it no longer zeroes the row's contribution.  The exponential, log-logistic,
and log-normal likelihoods do not use `time_lower` as an entry time and are
unaffected.
:::

Fit a Weibull model using both censoring types:

```{r}
#| label: interval-fit
fit_ic <- hazard(
  time       = time,
  status     = status,
  time_lower = time_lower,
  time_upper = time_upper,
  dist  = "weibull",
  theta = c(mu = 0.025, nu = 1.0),
  fit   = TRUE
)

fit_ic
```

### Interval-censored vs naive right-censored

A naive analyst who doesn't have visit-bracket data would record the death
at the *discovery* visit (`time_upper`) rather than as an interval.
This overstates the event time — the patient appears to have survived
longer than they did — biasing the estimated hazard shape.

```{r}
#| label: interval-vs-naive
# Naive: treat each interval-censored row as an exact event at time_upper
# (the visit when the death was first recorded)
fit_naive <- hazard(
  time   = ifelse(status == 2L, time_upper, time),
  status = ifelse(status == 2L, 1L,         status),
  dist  = "weibull",
  theta = c(mu = 0.025, nu = 1.0),
  fit   = TRUE
)

# Compare MLEs; true parameters are mu = 0.025, nu = 1.0
rbind(
  interval_censored = round(coef(fit_ic)[1:2],   4),
  naive_exact_upper = round(coef(fit_naive)[1:2], 4),
  truth             = c(mu = 0.025, nu = 1.0)
)
```

The interval-censored fit recovers both parameters accurately.  The naive
fit estimates $\mu$ comparably but introduces a substantial positive bias
in $\nu$ — it sees events consistently appearing at visit times (every 6
months) and infers a sharper, more periodic hazard shape that doesn't
match the underlying exponential structure.


## Convergence troubleshooting

Multiphase models have more parameters than single-distribution fits, and the
likelihood surface can be flat or multimodal when the data doesn't strongly
identify all of them. Most convergence problems trace back to one of three
causes: poor starting values, too many free parameters for the data at hand,
or a model that asks for a phase the data doesn't contain. This section shows
how to diagnose each and what to do about it.

### Reading the KM cumulative hazard for starting values

For a Weibull model, the relationship between the starting values and the
data is direct: because $H(t) = (\mu t)^\nu$, taking logs gives
$\log H(t) = \nu \log t + \nu \log \mu$. Plot $\log H(t)$ (the Nelson-Aalen
cumulative hazard on the log scale) against $\log t$ and fit a straight line;
the slope is $\hat\nu$ and the intercept is $\hat\nu \log \hat\mu$.

```{r}
#| label: starting-values
nel <- hzr_nelson(cabgkul$int_dead, cabgkul$dead)

# Drop the zero-time boundary point before log-transforming
nel_clean <- nel[nel$time > 0 & nel$cumhaz > 0, ]
log_t <- log(nel_clean$time)
log_H <- log(nel_clean$cumhaz)

lm_fit  <- lm(log_H ~ log_t)
nu_hat  <- unname(coef(lm_fit)[2])
mu_hat  <- exp(unname(coef(lm_fit)[1]) / nu_hat)
c(mu = round(mu_hat, 4), nu = round(nu_hat, 4))
```

Those values are the Weibull starting point the data itself suggests. They
won't be the MLEs --- the log-log line uses every event time equally, whereas
the MLE weights by the likelihood --- but they land the optimizer in a sensible
neighbourhood and prevent false-convergence to a degenerate solution.

For multiphase models the reading is qualitative. Plot the KM cumulative hazard
on the natural scale and look for kinks: a steep-then-flattening shape in the
first few months signals an early phase whose `t_half` should sit in that early
window; a steady linear rise afterwards signals a constant phase; a late
upward curve beyond the flat region signals a late phase. Set `t_half` to
roughly the time at which the early kink levels off, and set the constant
phase scale mu to approximate the slope of the linear mid-section.

```{r}
#| label: fig-cumhaz-shape
#| fig-cap: "Nelson-Aalen cumulative hazard for CABGKUL — three-phase shape visible as two kinks"
#| fig-width: 7
#| fig-height: 4
ggplot(data.frame(time = nel$time, cumhaz = nel$cumhaz),
       aes(time, cumhaz)) +
  geom_step(colour = "#D55E00", linewidth = 0.7) +
  labs(x = "Months after CABG", y = "Cumulative hazard") +
  theme_minimal()
```

The early steep rise, the mid-range roughly-linear section, and the late
upward acceleration map directly onto the three phases in the CABGKUL model.

### When to fix shape parameters

`fixed = "shapes"` locks the temporal shape parameters (e.g., `t_half`,
`nu`, `m` for a CDF phase; `tau`, `gamma`, `alpha`, `eta` for a g3 phase)
and lets the optimizer estimate only the scale (`log_mu`) for each phase.
This cuts the parameter count substantially and is almost always what you
want in two situations:

- **You have a reference fit.** If a SAS HAZARD run already produced shape
  estimates, replicate those shapes exactly and re-estimate only the scale.
  This is the standard parity workflow.
- **Your sample is small relative to the number of phases.** A rough guide
  is 50 or more events per free shape parameter. Below that threshold the
  shapes are weakly identified and the optimizer wanders.

Estimate shapes freely (`fixed = "none"`, the default) only when you are
exploring a new dataset for the first time and have no prior on the temporal
structure — and even then, start with `fixed = "shapes"` and release the
shapes one at a time if the fixed-shapes fit shows systematic misfit.

### Signs of overparameterization

When a model has more phases than the data supports, the symptoms are
recognizable:

| Symptom | What it means |
|---------|--------------|
| `fit$fit$converged == FALSE` | Optimizer hit `maxit` without satisfying the gradient tolerance |
| `vcov()` returns `NA` | Hessian is singular — parameters are not jointly identified |
| A phase scale (`log_mu`) at a boundary value | That phase is contributing essentially zero hazard; it isn't needed |
| Enormous standard errors on one or more parameters | Flat likelihood in that direction — weak identification |
| Two phases with nearly identical shapes | One can be collapsed into the other |

The AVC dataset has a clear two-phase structure (early CDF plus constant).
Adding a third late-rising phase asks the data for a pattern it doesn't
contain over this follow-up window:

```{r}
#| label: overparameterized
set.seed(42)
fit_3ph <- hazard(
  Surv(int_dead, dead) ~ 1,
  data   = avc,
  dist   = "multiphase",
  phases = list(
    early    = hzr_phase("cdf", t_half = 0.5, nu = 1, m = 1,
                          fixed = "shapes"),
    constant = hzr_phase("constant"),
    late     = hzr_phase("g3", tau = 5, gamma = 3, alpha = 1, eta = 1,
                          fixed = "shapes")
  ),
  fit = TRUE, control = list(n_starts = 5, maxit = 1000)
)

# Scale magnitudes: exp(log_mu) ≈ 0 for any phase the data doesn't support
log_mu_idx <- grep("log_mu", names(coef(fit_3ph)))
round(exp(coef(fit_3ph)[log_mu_idx]), 6)
```

The constant and late phase scales are both near zero — the optimizer
found no evidence in the AVC data for either of those temporal shapes over
this follow-up window. The early phase is absorbing all the identifiable
structure. Any phase whose fitted scale is $\mu \lesssim 10^{-4}$ is not
contributing meaningful hazard and is a candidate for removal. The
right diagnostic question is not "did the AIC improve?" but "does this
phase represent real clinical biology?" — late deterioration after AVC
repair requires long follow-up to observe; this dataset doesn't have it.

When shapes are also free (`fixed = "none"`), a redundant phase can make
the Hessian rank-deficient and `vcov()` will return `NA` (the Hessian
inversion failed). Fix by adding `fixed = "shapes"` to each phase to
reduce the free-parameter count, then release shapes one at a time if the
fixed-shapes fit shows systematic misfit.

### Optimizer options

Three `control` arguments address most remaining convergence problems:

**`n_starts`** (default 5 for multiphase; single-distribution models run
one optimizer call and ignore this parameter) runs the optimizer from
`n_starts` randomly jittered copies of the initial `theta`, then keeps the
best solution. The default of 5 is sufficient for two-phase models; increase
to 8–10 for three or more phases where the likelihood surface has more local
minima.

```{r}
#| label: nstarts
set.seed(42)
fit_robust <- hazard(
  Surv(int_dead, dead) ~ 1,
  data   = avc,
  dist   = "multiphase",
  phases = list(
    early    = hzr_phase("cdf", t_half = 0.5, nu = 1, m = 1,
                          fixed = "shapes"),
    constant = hzr_phase("constant")
  ),
  fit     = TRUE,
  control = list(n_starts = 10, maxit = 1000)
)
fit_robust$fit$converged
```

**Optimizer strategy (automatic).** The package always uses BFGS as the
primary optimizer. For fixed-shape multiphase models with 2–10 free
parameters, a Nelder-Mead pass runs first on each start to find a good
basin, then BFGS polishes the solution. This warm-up is transparent —
it happens automatically when the conditions are met and there is no
user-facing control to switch it on or off.

**`maxit`** (default 1000) sets the BFGS iteration cap. Hitting `maxit`
without converging usually means either the starting values are far from
the optimum (fix with better starts or `n_starts`) or the model is
overparameterized (fix by fixing shapes or dropping a phase). Raising
`maxit` beyond 2000 rarely helps if the optimizer is genuinely stuck.


## Phase types reference

You've now seen each phase type in use: a `"cdf"` early phase for AVC
operative mortality, a `"constant"` phase for AVC background rate, and
the implicit single shape of every Weibull fit. The package supports
three phase types in total, summarized here for quick reference:

| Type | Description | Typical use |
|------|-------------|-------------|
| `"cdf"` | Sigmoidal CDF shape (parameterized by `t_half`, `nu`, `m`) | Early or late phases with transient risk |
| `"constant"` | Flat hazard (no temporal shape parameters) | Ongoing background risk |
| `"g3"` | Late-phase G3 parameterization (4 parameters: `tau`, `gamma`, `alpha`, `eta`) | Late-rising risk matching C/SAS G3 output |

The `"cdf"` type covers the widest range of shapes: setting `t_half`
small (e.g., 0.5) creates an early-peaking phase; setting it large
(e.g., 10) creates a late-rising phase. The `"constant"` phase needs no
shape parameters. The `"g3"` shape is the explicit late-rising parameterization
that matches the SAS HAZARD "late" library; use it when you need
parity against a C/SAS reference fit, or when the late rise has a
clear lag-then-accelerate pattern that a delayed `"cdf"` doesn't
capture cleanly. See `vignette("mf-mathematical-foundations")` for the
full mathematical treatment of each, including the parameter
identifiability constraints.
