---
title: "Getting Started with TemporalHazard"
subtitle: "Build, fit, and predict with temporal parametric hazard models"
format:
  html:
    code-fold: false
vignette: >
  %\VignetteIndexEntry{Getting Started with TemporalHazard}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
has_pkg    <- requireNamespace("TemporalHazard", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2",       quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_pkg
)
```

TemporalHazard fits *parametric hazard models* to time-to-event data.
If you've worked with `survival::coxph()` or Kaplan-Meier curves before
you already have most of the vocabulary; this is a quick refresh of the
bits that matter for what follows.

A **hazard** is the instantaneous risk of the event at time $t$, given
the subject has survived until $t$. A Kaplan-Meier curve reflects it
implicitly: KM gives you survival probabilities, and the rate at which
those probabilities drop is (minus) the hazard. A **parametric** hazard
model writes the hazard as a smooth function of $t$ with a small number
of parameters, plus covariate effects. The payoff over nonparametric
KM is a smooth curve you can extrapolate, patient-specific risk
predictions, and the ability to compare fitted hazard *shapes* across
groups or models.

This vignette walks the minimal workflow. We start with a
single-distribution Weibull fit on simulated data (fit, summary,
predict), then move to the multiphase fit on CABGKUL, the canonical
three-phase cardiac-surgery example.

## A first Weibull fit

The simulated data below has 180 patients with three plausible risk
covariates: age, NYHA functional class, and cardiogenic shock. We fit
a single Weibull hazard: a scale parameter (`mu`) and a shape exponent
(`nu`) that lets the hazard accelerate when `nu > 1`, decelerate when
`nu < 1`, or stay flat at `nu = 1`. The Weibull is the natural starting
point for a single-distribution fit because it covers all three
monotone shapes with just two parameters.

The `hazard()` interface takes a survival formula
(`Surv(time, status)` on the left, covariates on the right), a
starting-value vector (`theta`), and a distribution name. Passing
`fit = TRUE` runs the optimizer and returns a fitted model; passing
`fit = FALSE` returns the model specification without fitting, which is
useful for sanity-checking the formula and starting values before
committing to the optimizer.

```{r}
library(TemporalHazard)
library(survival)

set.seed(1001)
n <- 180
dat <- data.frame(
  time = rexp(n, rate = 0.35) + 0.05,
  status = rbinom(n, size = 1, prob = 0.6),
  age = rnorm(n, mean = 62, sd = 11),
  nyha = sample(1:4, n, replace = TRUE),
  shock = rbinom(n, size = 1, prob = 0.18)
)

fit <- TemporalHazard::hazard(
  Surv(time, status) ~ age + nyha + shock,
  data = dat,
  theta = c(mu = 0.25, nu = 1.10, beta1 = 0, beta2 = 0, beta3 = 0),
  dist = "weibull",
  fit = TRUE,
  control = list(maxit = 300)
)

summary(fit)
```

## Prediction workflow

A fitted hazard model lets you score new patients without refitting.
The `predict()` method takes a `newdata` frame and a `type` argument
that selects which quantity to compute:

- `"linear_predictor"`: the covariate-driven log-hazard-ratio for
  each row, $x^\top \beta$.
- `"hazard"`: the hazard multiplier, $\exp(x^\top \beta)$.
- `"survival"`: the probability of surviving past each row's `time`,
  $S(t \mid x)$.
- `"cumulative_hazard"`: the integrated hazard up to each row's
  `time`, $H(t \mid x) = -\log S(t \mid x)$.

The three rows below are made-up patients spanning the covariate range
in the training data: a younger, low-NYHA, no-shock patient at an
early time; a middle case at the median; and an older, high-NYHA,
shock patient further out. We compute all four prediction types for
each.

```{r}
new_patients <- data.frame(
  time = c(0.5, 1.5, 3.0),
  age = c(50, 65, 75),
  nyha = c(1, 3, 4),
  shock = c(0, 0, 1)
)

pred_input <- new_patients

new_patients$linear_predictor <- predict(fit, newdata = pred_input, type = "linear_predictor")
new_patients$hazard_multiplier <- predict(fit, newdata = pred_input, type = "hazard")
new_patients$survival <- predict(fit, newdata = pred_input, type = "survival")
new_patients$cumulative_hazard <- predict(fit, newdata = pred_input, type = "cumulative_hazard")

new_patients
```

## Visualizing predicted survival

The summary table and the patient-level scores tell us the fit
*converged* and what it *predicts*, but not whether the predicted shape
matches the data. For that we plot the model. Below we build a smooth
survival curve over a fine time grid for a median-profile patient and
overlay the Kaplan-Meier nonparametric estimate from the raw data on
the same axes. If the parametric curve hugs the KM step function the
Weibull shape is honest to the data; visible drift in either direction
flags a model the optimizer fit happily but that doesn't actually
describe the cohort.

```{r}
#| label: fig-survival
#| fig-cap: "Parametric Weibull survival curve with Kaplan-Meier overlay"
#| fig-width: 7
#| fig-height: 5
#| eval: !expr requireNamespace("TemporalHazard", quietly = TRUE) && requireNamespace("ggplot2", quietly = TRUE)
library(ggplot2)

# Parametric curve on a fine grid
t_grid <- seq(0.05, max(dat$time), length.out = 80)
curve_df <- data.frame(
  time  = t_grid,
  age   = median(dat$age),
  nyha  = 2,
  shock = 0
)
curve_df$survival <- predict(fit, newdata = curve_df, type = "survival") * 100

# Kaplan-Meier empirical overlay
km <- survival::survfit(survival::Surv(time, status) ~ 1, data = dat)
km_df <- data.frame(time = km$time, survival = km$surv * 100)

ggplot() +
  geom_step(data = km_df, aes(time, survival, colour = "Kaplan-Meier")) +
  geom_line(data = curve_df, aes(time, survival,
                                  colour = "Parametric (Weibull)")) +
  scale_colour_manual(
    values = c("Parametric (Weibull)" = "#0072B2",
               "Kaplan-Meier"         = "#D55E00")
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(x = "Months after surgery", y = "Freedom from death (%)",
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Multiphase models

The Weibull fit above describes the hazard with one smooth, monotone
curve. It can rise, fall, or stay flat over time, but it cannot change
direction. That's fine for some processes (light bulbs failing, parts
wearing out). It breaks down quickly for clinical data.

Consider CABG (coronary artery bypass graft) surgery. In the days right
after the operation the risk of death is at its highest: the patient is
in the ICU, exposed to operative complications, infections, bleeding.
If they survive that window the risk drops to a low, roughly flat
background rate: they recover, go home, live their lives. Years later
the risk rises again as the grafts begin to deteriorate and the cohort
ages. Three distinct regimes, each with its own clinical mechanism.

No single Weibull (or log-normal, or any other monotone distribution)
can fit all three at once. The optimizer would have to compromise:
picking a shape that's mediocre everywhere instead of right in any one
regime. You'd predict too few early deaths, the wrong slope in the
middle, and too few late deaths.

The **multiphase decomposition** is the workaround. Instead of asking
one distribution to do everything, we describe the total cumulative
hazard as a *sum* of phase-specific contributions, each with its own
temporal shape:

$$H(t \mid x) = \sum_{j=1}^{J} \mu_j(x) \cdot \Phi_j(t)$$

Each $\Phi_j(t)$ is a phase-specific *temporal shape*: a unit-scaled
curve that says *how* phase $j$'s hazard accumulates over time. Each
$\mu_j(x)$ is the phase-specific *scale*: how much cumulative hazard
belongs to phase $j$, possibly modulated by covariates. The phases
overlap and their hazards add. There's no switching, no thresholds, no
piecewise joins; at any time $t$ the instantaneous hazard $h(t) = dH/dt$
is the sum of the per-phase contributions $\mu_j \cdot \phi_j(t)$, where
$\phi_j = d\Phi_j/dt$.

### Why multiple phases?

Because the number of phases is the modeling lever that lets you match
the data's actual structure. Some processes are well-described by two
phases (early failure + steady-state). Most cardiac-surgery cohorts
need three: early operative risk + constant background + late
deterioration. A small handful need four (operative + subacute +
constant + late); acute aortic dissection is the classic. The
framework is general; the *count* is a modeling choice driven by
clinical knowledge and the shape of the Kaplan-Meier cumulative hazard
curve.

For the CABGKUL fit below we use three phases. That's not a property of
the package; it's a property of this dataset.

Each phase is specified with `hzr_phase()`. The first argument picks
the shape function: `"cdf"` for a saturating curve bounded between 0
and 1 (the SAS "early" / G1 shape), `"constant"` for a flat hazard
plateau (SAS "G2"), `"g3"` for the power-law late-rising shape from
the SAS "late" library. Remaining arguments set the shape's free
parameters, or fix them with `fixed = "shapes"`. The Phase types
section below has the full menu.

For this fit we use the textbook three-phase decomposition: a
saturating early peak, a constant background, and a power-law late
rise. We fix the shapes so we can compare the fitted scales against
the published reference; in your own work you would usually estimate
them.

```{r}
#| label: multiphase-fit
# CABGKUL is the benchmark dataset for 3-phase decomposition (n = 5,880)
data(cabgkul)

fit_mp <- hazard(
  Surv(int_dead, dead) ~ 1,
  data   = cabgkul,
  dist   = "multiphase",
  phases = list(
    early    = hzr_phase("cdf", t_half = 0.2, nu = 1, m = 1,
                          fixed = "shapes"),
    constant = hzr_phase("constant"),
    late     = hzr_phase("g3",  tau = 1, gamma = 3, alpha = 1, eta = 1,
                          fixed = "shapes")
  ),
  fit     = TRUE,
  control = list(n_starts = 5, maxit = 1000)
)

summary(fit_mp)
```

The summary marks most rows `NA` in the `std_error`, `z_stat`, and
`p_value` columns. Two mechanisms produce these, both deliberate:

- The seven shape parameters carry `fixed = "shapes"` and act as
  user-supplied constants (`t_half`, `nu`, `m` on the early phase; `tau`,
  `gamma`, `alpha`, `eta` on the late phase). The optimizer never moves
  them, so they have no estimated variance to report.
- One of the three `log_mu` scale parameters is also `NA`. By default the
  multiphase optimizer enforces Conservation of Events (Turner's theorem):
  one phase's `log_mu` is solved analytically at every iteration so that
  total predicted events equal total observed. That phase is not a free
  parameter at the MLE, so its inverse-Hessian variance is degenerate, and
  the table marks it `NA` rather than print a misleading number. Pass
  `control = list(conserve = FALSE)` to disable Conservation of Events and
  estimate all three scale parameters freely.

The two remaining `log_mu` rows are the optimizer-estimated free
parameters; their Wald statistics reflect the two-parameter fit
dimension that remains after the Conservation-of-Events constraint
absorbs the third `log_mu`.

### Decomposed hazard visualization

The MLE numbers in the summary tell us *what* the model committed to,
but not *how* the three phases fit together over time. For that we
need the per-phase contributions. The `decompose = TRUE` argument on
`predict()` returns one cumulative-hazard column per phase plus the
total; we numerically differentiate each one (a coarse first-difference
is fine here, since we just want to look) to get an instantaneous
hazard rate.

The plot that follows shows whether
the model carved up the timeline the way we expected. The early phase
should dominate near $t = 0$ and die off. The constant phase should be
a flat floor. The late phase should be near zero early and rise after
a lag. If any of those shapes looks wrong, the fix is in the starting
values or in the choice of shape function, not in the data.

```{r}
#| label: fig-hazard-phases
#| fig-cap: "Additive phase decomposition: total hazard (solid) = early + constant + late (dashed)"
#| fig-width: 7
#| fig-height: 4.5
#| eval: !expr requireNamespace("TemporalHazard", quietly = TRUE) && requireNamespace("ggplot2", quietly = TRUE)
t_grid <- seq(0.01, max(cabgkul$int_dead) * 0.95, length.out = 200)
nd     <- data.frame(time = t_grid)

# decompose = TRUE returns per-phase cumulative hazard columns
decomp <- predict(fit_mp, newdata = nd, type = "cumulative_hazard",
                  decompose = TRUE)

# Numerical differentiation: h(t) ≈ ΔH(t) / Δt
num_hazard <- function(cumhaz, time) {
  dt <- diff(time)
  dH <- diff(cumhaz)
  c(dH[1] / dt[1], dH / dt)
}

h_long <- rbind(
  data.frame(time = t_grid, hazard = num_hazard(decomp$early, t_grid),
             Phase = "Early"),
  data.frame(time = t_grid, hazard = num_hazard(decomp$constant, t_grid),
             Phase = "Constant"),
  data.frame(time = t_grid, hazard = num_hazard(decomp$late, t_grid),
             Phase = "Late"),
  data.frame(time = t_grid, hazard = num_hazard(decomp$total, t_grid),
             Phase = "Total")
)
h_long$Phase <- factor(h_long$Phase,
                       levels = c("Total", "Early", "Constant", "Late"))

ggplot(h_long, aes(time, hazard, colour = Phase, linetype = Phase)) +
  geom_line(aes(linewidth = Phase)) +
  scale_colour_manual(values = c(Total = "#222222", Early = "#E69F00",
                                 Constant = "#56B4E9", Late = "#CC79A7")) +
  scale_linetype_manual(values = c(Total = "solid", Early = "dashed",
                                   Constant = "dashed", Late = "dashed")) +
  scale_linewidth_manual(values = c(Total = 1.3, Early = 0.7,
                                    Constant = 0.7, Late = 0.7)) +
  labs(x = "Months after CABG", y = "Hazard rate",
       colour = "Phase", linetype = "Phase", linewidth = "Phase") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Multiphase survival with Kaplan-Meier overlay

The decomposition plot tells us the model has the right *shape*; the
overlay plot tells us it has the right *level*. Plotting the
parametric survival curve against the Kaplan-Meier estimate on the
same axis is the single most useful diagnostic for a hazard fit. KM
is the gold-standard nonparametric reference; if the parametric curve
drifts away from it, something is wrong with the model that more
iterations won't fix.

```{r}
#| label: fig-mp-survival
#| fig-cap: "Multiphase parametric survival vs. Kaplan-Meier"
#| fig-width: 7
#| fig-height: 4.5
#| eval: !expr requireNamespace("TemporalHazard", quietly = TRUE) && requireNamespace("ggplot2", quietly = TRUE)
surv_df <- data.frame(
  time     = t_grid,
  survival = predict(fit_mp, newdata = nd, type = "survival") * 100
)

km    <- survfit(Surv(int_dead, dead) ~ 1, data = cabgkul)
km_df <- data.frame(time = km$time, survival = km$surv * 100)

ggplot() +
  geom_step(data = km_df, aes(time, survival, colour = "Kaplan-Meier"),
            linewidth = 0.6) +
  geom_line(data = surv_df, aes(time, survival, colour = "Multiphase model"),
            linewidth = 1.1) +
  scale_colour_manual(
    values = c("Multiphase model" = "#0072B2", "Kaplan-Meier" = "#D55E00")
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(x = "Months after CABG", y = "Freedom from death (%)",
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Phase types

Picking the right phase shape is the single biggest modeling decision
in a multiphase fit. The package ships the canonical shapes from the
SAS C HAZARD library:

- **`"cdf"`**: a saturating curve $\Phi(t) \in [0, 1]$, parameterized
  by `t_half` (the time at which $G(t_{1/2}) = 0.5$), `nu` (a time
  exponent), and `m` (a shape exponent). This is the SAS "early" or G1
  shape. Use it for the early phase, or for any phase whose accumulated
  hazard saturates rather than growing without bound.
- **`"constant"`**: $\Phi(t) = t$. The phase's hazard rate is flat;
  there are no shape parameters to estimate, only the scale $\mu$. This
  is the SAS G2 shape. Use it for a background plateau.
- **`"g3"`**: the SAS "late" shape, parameterized by `tau` (a
  time-scale), `gamma` (a time exponent), `alpha` (a shape parameter;
  $\alpha = 0$ gives an exponential limit), and `eta` (an outer
  exponent). Use it for a late phase that is near-zero early and
  accelerates over time; graft deterioration over years is the
  archetype.

See `?hzr_phase` for the additional `"hazard"` shape and the full
parameterization of each.

In practice you choose phase shapes by looking at the Kaplan-Meier
cumulative hazard plot, identifying which regimes are present, and
matching each regime to the shape family whose behavior fits. Then
you fit, look at the decomposition plot, and adjust starting values
or fix shapes if a phase doesn't land where you expected.

```{r}
#| eval: false
hzr_phase("cdf",      t_half = 0.5, nu = 2, m = 1)   # Early risk (bounded)
hzr_phase("constant")                                   # Flat background rate
hzr_phase("cdf",      t_half = 10,  nu = 1, m = 0)    # Late risk (CDF-based)
hzr_phase("g3",       tau = 1, gamma = 3, alpha = 1,   # Late risk (G3 power law)
                       eta = 1)
```

The `"g3"` type uses the four-parameter G3 decomposition from the original
C/SAS HAZARD program, providing unbounded power-law growth for late-phase
hazards (exponential growth in the `alpha = 0` limit). See `vignette("mf-mathematical-foundations")` for the full
mathematical treatment.

## Numerical helpers

The package ships a small set of numerical helpers used internally to
keep the likelihood evaluations stable across the wide range of
arguments an optimizer encounters. They're exported so you can reach
for them in custom code or when debugging a fit.

The most common one is `hzr_log1pexp(x)`, which computes
$\log(1 + e^x)$ in a numerically stable way. The naive `log(1 + exp(x))`
overflows once $x$ exceeds about 700 (because `exp(x)` itself
overflows); the helper switches to the asymptotic $x + \log(1 + e^{-x})$
for large $x$, which is exact to floating-point precision. The package
uses it in `hzr_decompos()`, the shape function behind the `"cdf"` and
`"hazard"` phase types, so the same evaluation can run on $x = -50$ and
$x = 5000$ without producing `Inf` or `NaN`.

```{r}
TemporalHazard::hzr_log1pexp(c(-2, 0, 2))
```
